---
title: "Lecture 3 Exercises"
author: "Isabel Fulcher"
date: "8/7/2018"
output: pdf_document
---

```{r,include=FALSE}
library("knitr")
```

# Normal distribution exercises

1. Generate 10 draws from a normal random variable with mean 5 and variance 4. What is the sample mean of these variables? What is the true mean? 
```{r} 
```

2. How would you expect the previous answer to change if we increased the number of draws to 10,000? Check your intuition.
```{r} 
```

3. The probability density function (pdf) for a normal random variable $X$ with mean $\mu$ and variance $\sigma^2$ is as follows, 

$$f_X(x) = \frac{1}{\sqrt{2\pi}\sigma} exp(\frac{1}{2\sigma^2}(x - \mu)^2) $$

What is the value of $f_X(0)$ when $\mu=0$ and $\sigma^2 = 2$? Calculate using the above expression and built-in R function. 

```{r} 
```

4. If $X \sim N(5,4)$, what is $P(X < 2)$? What is $P(X \leq 2)$? What is $P(X > 2)$?
```{r} 
```


# Poisson distribution exercises
1. Generate a random sample of size 100,000 from a Poisson distribution with rate parameter 5
```{r} 
```
2. Compute the sample mean and variance. What do you notice?
```{r} 
```
3. Compute the empirical estimates of $P(X = 0)$, $P (X = 1)$, and $P(X = 2)$ for $X \sim Pois(5)$ from your sample. 
```{r} 
```
4. Compare the estimates to the true value of the probability mass function (pmf) at $x=0,1,2$. Use the pmf and the built-in R function. Note the pmf is given by: 

$$ P(X=x) = \frac{\lambda^x e^{-\lambda}}{x!} \textrm{ where } x \in 0,1,2,... \textrm{ and } \lambda > 0$$

```{r} 
```

5. Compute an estimate of $P(0 < X < 3)$ from the sample and compare to true value. 
```{r} 
```

6. Find the smallest $k$ such that $0.30 < P(X < k)$ for $X \sim Pois(5)$ using a while loop. Then, do this using a built-in R function. 
```{r}
```


\newpage

# Group exercises
To display your results, create a table in Rmarkdown using the kable() function. Try to make it as clean as possible (i.e. column headers, title, digits, etc.). 

Fun note: you will prove these results formally in the probability course! 

## Group 1 
Generate 100,000 samples from a geometric distribution with $p = 0.3$.  Estimate $P(X \ge s+t | X \ge t)$ and $P(X \ge s)$ for  $s = 4$ and $t = 1, 2, 3, 4, 5, 6$. Compare to the true values.  What do you notice?  Google `memoryless property distribution' and take a look at the wiki page on memorylessness.  What does this suggest about the geometric distribution? 

```{r} 
```

## Group 2
Generate 10,000 samples from a $Bin(3, 0.5)$ and another 10,000 samples from $Bin(5, 0.5)$.  Compute the empirical cdf of the sum of the two samples and compare to the distribution function of a $Z \sim Bin(8,0.5)$ random variable.  What does this suggest about the distribution of $X+Y$ where $X \sim Bin(n_1, p)$ and $Y \sim Bin(n_2, p)$? 

```{r} 
```


