---
title: "Lecture 7 Exercises"
author: "Isabel Fulcher"
date: "8/16/2018"
output: pdf_document
urlcolor: blue
---

## Install packages 
```{r,message=FALSE}
library(matrixStats)
library(knitr)
library(tidyverse)
library(reshape2)
library(MASS)
```

Load in the infants dataset from Lecture 5. We are again interested in the relationship between birthweight $Y$, smoking $X_1$, and mother's weight $X_2$. 
```{r}
load("infants.dat")
```

# Exercise 1

Recall, the likelihood for a linear model where we assume $\epsilon_i \sim N(0,\sigma^2)$ and observe $X_1, ... X_n$ is, 

$$ \mathcal{L}(\beta_0,\beta_1,\beta_2) = \prod_{i=1}^n \frac{1}{\sqrt{2\pi}\sigma} exp( \frac{-1}{2\sigma^2} (Y_i - (\beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2}))^2 )$$
The log-likelihood can then be written as, 

$$ \ell(\beta_0,\beta_1,\beta_2) = \sum_{i=1}^n -\log(\sqrt{2\pi}\sigma) - \frac{1}{2\sigma^2} (Y_i - (\beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2}))^2$$

1. Write a function that calculates the negative log-likelihood and takes in values for $Y$, $X_1$, and $X_2$, which are all vectors of length n, and a vector for the unknown parameters, i.e. $\{ \beta_0, \beta_1, \beta_2, \sigma^2 \}$.  
```{r}
loglik <- function(par,Y,X){
  beta <- par[1:3]
  sigma2 <- par[4]
  l <- sum(log(sqrt(2*pi*sigma2)) + (1/(2*sigma2))*(Y-X%*%beta)^2)
  return(l)
}
```


2. Use the optim() function to find the the MLE of $\boldsymbol{\beta}$ when the outcome $Y$ is birthweight, $X1$ is smoking, and $X2$ mother's weight. NOTE: you would not typically do this in practice because there is a closed-form solution (recall OLS estimates!). This is just for illustration.
```{r,message=FALSE,error=FALSE,warning=FALSE,cache=TRUE}
X <- cbind(1,infants$smoker,infants$gained)
Y <- infants$weight

fit.optim <- optim(runif(4,0,1),loglik,Y=Y,X=X,method="BFGS")

beta.optim <- fit.optim$par[1:3]
```

3. Calculate the OLS estimate for $\boldsymbol{\beta}$ using R and the analytical expression give in Lecture 5. How does this compare to the above? 
```{r, results='hide'}
#Using lm() function
fit.ols1 <- lm(weight ~ smoker + gained, data=infants)
beta.ols1 <- coefficients(fit.ols1)

#Using actual values
beta.ols2 <- t(solve(crossprod(X))%*%t(X)%*%Y)

#Compare all methods
beta.ols1 ; beta.ols2 ; beta.optim
```


# Exercise 2: Logistic regression

A logistic regression model is given by,

$$ \textrm{logit}(Pr(Y=1 | X_1, X_2)) = \alpha_0 + \alpha_1X_1 + \alpha_2 X_2 \implies Pr(Y=1 | X_1, X_2) = \textrm{expit}(\alpha_0 + \alpha_1X_1 + \alpha_2 X_2) $$

The likelihood for a logistic model where we observe $X_1, ... X_n$ is given by, 

$$ \mathcal{L}(\alpha_0,\alpha_1,\alpha_2) = \prod_{i=1}^n Pr(Y_i=1 | X_{i1}, X_{i2})^{Y_i}(1-Pr(Y_i=1 | X_{i1}, X_{i2}))^{1-Y_i} $$

The log-likelihood can be written as, 

$$ \ell(\alpha_0,\alpha_1,\alpha_2) = \sum_{i=1}^n Y_i(\alpha_0 + \alpha_1 X_{i1} + \alpha_2 X_{i2}) - \log[1+\exp(\alpha_0 + \alpha_1 X_{i1} + \alpha_2 X_{i2})]$$

1. Write a function for that calculates the negative log-likelihood. This function should take in values for the data, i.e. $Y$, $X_1$, and $X_2$, which are all vectors of length $n$, and $\boldsymbol{\alpha}=(\alpha_0,\alpha_1,\alpha_2)$. 

```{r}
infants %>% mutate(weight.binary = ifelse(weight <= 2.5,1,0)) -> infants # create new outcome

negloglik <- function(par,Y,X){
  alpha <- par[1:3]
  l <- sum(-(Y*(X%*%alpha)) + log(1 + exp(X%*%alpha)))
  return(l)
}

```

2. Use the optim() function to estimate the the MLE of $\boldsymbol{\alpha}$ in this dataset.
```{r}
X <- cbind(1,infants$smoker,infants$gained)
Y <- infants$weight.binary

fit.optim <- optim(runif(4,0,1),negloglik,Y=Y,X=X,method="BFGS")

beta.optim <- fit.optim$par[1:3]
```

3. Check your answer using the built-in R function for logistic regression (and estimation of parameters in GLMs in general). Use the glm() function, 
```{r, eval=FALSE}
# the glm() function has the following form 
# replace with your data, but keep the family argument
fit.glm <- glm(infants$weight.binary ~ infants$smoker + infants$gained, family=binomial) 
summary(fit.glm)
```


