---
title: "Lecture 2 Exercises"
author: "Isabel Fulcher"
date: "8/3/2018"
output: pdf_document
---

```{r,include=FALSE}
library("dplyr")
library("matrixStats")
```

# Part 1: Matrix / Vector Operations

Before starting generate the following in R:

```{r}
set.seed(11)
b <- sample(10,4,replace=TRUE)
A <- matrix(sample(10,16,replace=TRUE),4,4)
```

Now perform the following exercises to get familiar with matrix operations: 

1. What is the dimension of $\mathbf{A}$?
```{r}
```
2. Find the transpose of $\mathbf{b}$
```{r}
```
3. Are vectors in R column or row vectors by default?
4. Calculate the following in R: $\mathbf{A}^T\mathbf{A}$
```{r}
```
5. Repeat the previous exercise with a built-in R function
```{r}
```
6. Solve $\mathbf{b} = \mathbf{A}\mathbf{x}$ for $\mathbf{x}$
```{r}
```
7. Repeat the previous with a built-in R function.
```{r}
```
8. Create a $4x4$ matrix with diagonal elements equal to 5 and off-diagonals equal to 0.
```{r}
```

# Part II: Flow control

## For Loops

1. Using a for loop, generate a vector containing the natural numbers up to 10 and add 2 to each element. How can you do this without a loop?
```{r}
```

2. Load in the iris dataset using the R code provided below. Use a for loop to compute the standard deviation of the first four columns measuring sepal length, sepal width, pedal length, and pedal width respectively. 
```{r} 
# load dataset 
data(iris)
# view the first ten observations
head(iris) 
```

## Apply functions

1. For question 2 in the "For Loops" section, use \verb+apply+ to perform the same task. 
```{r} 
```


2. Report the 20th and 80th percentiles for each variable in the Iris dataset. Hint: use the \verb+quantile+ function. 
```{r} 
```

3. Compute the variance-covariance matrix for the four variables in the Iris dataset. Confirm your answer by using the \verb+cov+ function. Hint: the covariance matrix is given by $(n-1)^{-1} \mathbf{X}^{*T} \mathbf{X}^*$ where $\mathbf{X}^*$ contains mean centered values for each of the four variables.

```{r}
```


4. In a given library of RNA-seq reads, there are duplicate observations due to PCR amplification. However, you know that there are truly $N = 50,000$ unique reads. Your collaborator wants to know how many reads she should sequence ($K \geq N$) to get a good saturation of her library (i.e. capture as many unique reads as possible). Assume that sampling from the N unique reads occurs with replacement and equal probability. What is the expected number of unique reads for the values of $K$ between 50,000 and 200,000 (use increments of 10,000)? 

```{r}
set.seed(345)
```


## While Loops and conditional statements 
1. Roll a fair six sided die; if the roll is a prime number, print "prime"; if the roll is a composite number, print "composite"; otherwise print "1" 
```{r} 
#Hint: the %in% function shows if an element is in a vector 
2 %in% c(1,2,3)
```

2. The initial value of the number is 0. If the number is less than 10, then add a random number between 0 and 1 to it and print the resulting sum. Continue until the total sum is greater than 10.

```{r}
#Hint: the runif function returns a random number between 0 and 1
runif(n=1,min=0,max=1)
```


3. The Newton-Raphson method (or Newton Method) is a simple iterative technique for finding the zero $r$ of a real-valued function $f(x)$. First, we provide some initial guess $x_0$ for $r$.  The goal is then to obtain a better estimate $x_n$ of $r$ in $n$ iterations. Using the Newton-Raphson method, find the $5^{th}$ root of 7.


+ The algorithm can be summarized as follows. First, we can rewrite $r=x_0+h$ for some $h$ which measures how far the estimate initial guess $x_0$ is from the true zero $r$.  Assuming $h$ is small, we can use linear approximation to conclude that
$$0 =f(r) =f(x_0 +h) \approx f(x_0)+ hf'(x_0)$$
 It then  follows that
$$ h \approx  - f(x_0) / f'(x_0)$$
and therefore a better estimate of $r$ is given by 
$$x_1 = x_0 + h \approx x_0 - f(x_0)/ f'(x_0)$$
Continuing in this way, if $x_n$ is the current estimate of $r$ then $x_{n+1}$ is  given by
$$x_{n+1} \approx x_n - f(x_n)/ f'(x_n)$$

+ Hint: You may want to specify a small pre-specified tolerance level in place of the approximations. In some cases, it may also be useful to specify a maximum number of iterations. 


```{r}
```
