---
title: "Lecture 4 Exercises"
author: "Isabel Fulcher"
date: "8/9/2018"
output: pdf_document
urlcolor: blue
---

## Load packages 
```{r,message=FALSE}
library("tidyverse") #ggplot2, dplyr, etc.
library("reshape2") #need this for melt() 
library("car") #contains dataset
```

# Part I: Basic data analysis using tidyverse

Open the Blackmore dataset in R. This contains data on 138 teenaged girls hospitalized for eating disorders and 98 control subjects. There are four variables: subject id, age in years, hours per week of exercise, and group (control vs. eating-disorder). There are multiple observations per subject. 

```{r} 
data(Blackmore)   
head(Blackmore)
```

## Key commands

### Pick observations by their values with filter()
```{r,results='hide'}
Blackmore %>% filter(group=="patient") 

# You do not have to use piping, just makes it cleaner
filter(Blackmore,group=="patient")

# You can also just use base R
Blackmore[Blackmore$group=="patient",]
```

### Pick variables by their names with select()
```{r,results='hide'}
Blackmore %>% dplyr::select(subject,exercise)

#and in base R...
Blackmore[c("subject","exercise")]
```

### Create new variables with functions of existing variables with mutate()
```{r,results='hide'}
Blackmore %>% mutate(age_new = age^2) 

#and in base R...
Blackmore$age_new = Blackmore$age^2
```

### Collapse many values down to a single summary with summarise()
```{r,results='hide'}
Blackmore %>% summarise(age_mean = mean(age),age_sd = sd(age))

#and in base R...
mean(Blackmore$age)
sd(Blackmore$age)
```

### To only perform the summarise command on a subset use group_by()
```{r,results='hide'}
Blackmore %>% group_by(group) %>% summarise(age_mean = mean(age),age_sd = sd(age)) 

#and in base R...
mean(Blackmore[Blackmore$group=="patient",]$age)
mean(Blackmore[Blackmore$group=="patient",]$age)
sd(Blackmore[Blackmore$group=="control",]$age)
sd(Blackmore[Blackmore$group=="patient",]$age)
```

## Plotting with ggplot2 

### The basic idea 
```{r,eval=FALSE}
# Initialize the plot with data and aesthetics
p <- ggplot(data=NULL,mapping=aes(x=,y=,fill=,group=,color=)) 

# Add layers 
p <- p + geom_histogram()

# Make pretty
p <- p + theme_bw() + labs(title="",x="",y="")
```

### Example: Plot a histogram of hours per week of exercise across all observations. 

```{r} 
# Initialize the plot with data and aesthetics
p <- ggplot(Blackmore,aes(x=exercise))

# Add layer 
p1 <- p + geom_histogram() 

p2 <- p + geom_histogram(aes(y=..density..))

# Make pretty
p3 <- p2 + theme_bw()
```

### Saving graphics 
```{r,eval=FALSE}
# Setting your working directory 
setwd("your/file/path/here")

# Use ggsave
ggsave("plot_name.pdf",plot = p)
```

\newpage

# Part I: Exercises
1. Plot a line for each person that shows exercise at each age observation. Color the line by group.
```{r,fig.width=4, fig.height=3} 
```

2. Create a new data frame that only contains the subject id, group, and average exercise per person. 

```{r,fig.width=4, fig.height=3} 
```

3. Using the data frame from question 2, plot two histograms of mean hours per week of exercise by group in the same figure.

```{r,fig.width=4, fig.height=3}
```

4. Calculate the 95\% confidence interval for the mean exercise hours for each group. The confidence interval is given by, 

$$(\bar{X} - z_{.975} \sigma / \sqrt{n}, \bar{X} + z_{.975} \sigma / \sqrt{n})$$
To calculate $z_{.975}$, use a command we learned last class (hint: starts with a 'q' and ends with a 'norm'). You can use the estimated standard deviation in place of $\sigma$ (for those of you who recognize that this is naughty, I'm unconcerned in this case becase $n$ is large in each group). 

```{r}
```

5. Plot these two confidence intervals on the same plot using geom_errorbar() 

```{r,,fig.width=5, fig.height=4} 
```

\newpage

# Part II: Central Limit Theorem (CLT) exercise 

The goal here is to illustrate the Central Limit Theorem. The Central Limit Theorem states that the sample mean will converge in distribution to a normal distribution. Given $X_1,...X_n$ independent draws from some underlying distribution, 

$$ \frac{(\bar{X}_n - \mu)}{\sigma / \sqrt{n}} \to N(0, 1) $$

1. Draw one random sample of size 100 from a Poisson distribution with event rate $\lambda$ equal to 1.5. Use a histogram to display the empirical probability distribution. Include a vertical line indicating the sample mean using geom_vline(). 
```{r} 
set.seed(1567)
```

2. Draw 10,000 random samples of size 100 from a Poisson distribution with event rate $\lambda$ equal to 1.5. Calculate the quantity on the left hand side of the above equation for each sample. Note that the poisson distribution has mean $E[X] = \mu = \lambda$ and $Var[X] = \sigma^2 = \lambda$.  
```{r} 
```

3. Construct a histogram of the quantites from question 2.

```{r,fig.width=4, fig.height=3} 
```

4. On the figure from question 3, overlay the pdf of the normal distribution corresponding to the right hand side of the expression. Hint: use stat_function(fun=dnorm). 

```{r,fig.width=4, fig.height=3}
```

5. Now, repeat questions 2-4 using various sample sizes of $N=(5,10,25,50,100,1000)$. You will want to collect all your results in one data frame for plotting.

```{r}
```

6. Plot the results from question 5 in ONE figure (i.e. panel of six plots) using the facet_wrap() function. Depending on how you saved your results from 5, you may also need to use the melt() function to get your data frame into the appropriate format for ggplot2-ing.
```{r,fig.width=6, fig.height=5} 
```


7. Let's motivate how this relates to confidence intervals for the population mean! Calculate the confidence intevals for 100 random samples of size $n=100$ from a Poisson distribution with $\lambda$ equal to 1.5. Plot the confidence intervals and include a line for the true value of $\lambda$. How often is the true mean captured? Does this match expectations?

Tip: to make it easier to see which confidence intervals cover the true mean, you may want to sort or color the intervals.

```{r}
set.seed(98)
```


8. Take a moment and interpret the implications from this exercise. How should we interpret a confidence interval?







# Great references / more practice with tidyverse

* Hadley's R for Data Science book or [website](http://r4ds.had.co.nz/)
* Specifically, the ggplot2 [exercises](http://r4ds.had.co.nz/graphics-for-communication.html)
* Tidyverse [cheatsheet](https://s3.amazonaws.com/assets.datacamp.com/blog_assets/Tidyverse+Cheat+Sheet.pdf) by DataCamp

