---
title: "Lecture 8 Exercises"
author: "Isabel Fulcher"
date: "8/20/2018"
output: pdf_document
urlcolor: blue
---

## Install packages 
```{r,message=FALSE}
library(tidyverse)
library(reshape2)
library(MASS)
```

# Bootstrap exercise

We are going to use the bootstrap to get the standard errors from the previous function that you wrote, which likely looked something like,

```{r}
negloglik = function(alpha, X, Y) {
  return(-sum(
    Y*(X%*%alpha)
    -log(1 + exp(X %*% alpha))
  ) 
  ) 
}
```

Recall this function corresponds to the following logistic model, 
$$ \textrm{logit}(Pr(Y=1 | X_1, X_2)) = \alpha_0 + \alpha_1X_1 + \alpha_2 X_2  $$
Let's create a dataset for this exercise.

```{r} 
set.seed(12345)
n <- 100
Sigma <- matrix(c(2,.2,.2,3),2,2)
alpha <- c(.5,.4,-.2)
X <- cbind(1,mvrnorm(n,rep(0,2),Sigma))
Y <- rbinom(n,1,plogis( X%*%alpha))

data <- data.frame(cbind(Y,X[,2:3]))
colnames(data) <- c("Y","X1","X2")
```

\newpage 
1. Perform the bootstrap using $B=1,000$ replicates for $\hat{\boldsymbol{\alpha}} = (\hat{\alpha}_0, \hat{\alpha}_1, \hat{\alpha}_2)$ and save the estimates from each replicate.  
```{r}
B <- 1000

#Option 1: For loop
beta.boot <- matrix(NA,B,3)
for (b in 1:B){
  
  #take a boostrap sample 
  data.boot <- data[sample(n,n,replace=TRUE),]
  X.boot <- as.matrix(cbind(1,data.boot[,2:3]))
  
  #fit
  fit.optim <- optim(runif(3,0,1),negloglik,Y=data.boot$Y,X=X.boot,method="BFGS")
  beta.boot[b,] <- fit.optim$par
  
}

# Option 2: sapply function
bootfunc <- function(b,data,n){
  data.boot <- data[sample(n,n,replace=TRUE),]
  X.boot <- as.matrix(cbind(1,data.boot[,2:3]))
  fit.optim <- optim(runif(3,0,1),negloglik,Y=data.boot$Y,X=X.boot,method="BFGS")
  return(fit.optim$par)
}

beta.boot <- t(sapply(1:B,bootfunc,data=data,n=n))

# Option 3: tidyverse purrr?

```

2. Using the results from 1, calculate the bootstrap standard error and 95% confidence intervals. 
```{r}
beta.df <- data.frame(beta.boot)
beta.df %>% melt() %>% group_by(variable) %>% 
  summarise(sd = sd(value), ci_low = quantile(value,probs=.025),
            ci_up = quantile(value,probs=.975)) 
```


3. Compare the bootstrap standard error to that from the glm() function. 
```{r} 
fit.glm <- glm(Y ~ X[,2] + X[,3],data=data,family="binomial")
summary(fit.glm)
```

Note that you can also calculate the standard error for for $\hat{\boldsymbol{\alpha}}$ from the optim() function using standard likelihood theory. I have provided the code for this below. 
```{r, results='hide'} 
fit.optim <- optim(runif(3,0,1),negloglik,Y=data$Y,X=X,method="BFGS",hessian=TRUE)
hess.optim <- fit.optim$hessian
se.optim <- sqrt(diag(solve(hess.optim))) 

se.optim
```

