---
title: "Lecture 3 Solutions"
author: "Isabel Fulcher"
date: "8/7/2018"
output: pdf_document
---

```{r,include=FALSE}
library("knitr")
```

# Normal distribution exercises

1. Generate 10 draws from a normal random variable with mean 5 and variance 4. What is the sample mean of these variables? What is the true mean? 
```{r} 
set.seed(12) 
x <- rnorm(10,5,2)
mean(x)
```

2. How would you expect the previous answer to change if we increased the number of draws to 10,000? Check your intuition.
```{r} 
set.seed(12)
x <- rnorm(10000,5,2)
mean(x)
```

3. The probability density function (pdf) for a normal random variable $X$ with mean $\mu$ and variance $\sigma^2$ is as follows, 

$$f_X(x) = \frac{1}{\sqrt{2\pi}\sigma} exp(\frac{-1}{2\sigma^2}(x - \mu)^2) $$

What is the value of $f_X(0)$ when $\mu=0$ and $\sigma^2 = 2$? Calculate using the above expression and built-in R function. 

```{r} 
#evaluate expression
(1/sqrt(2*pi*2))

#built-in R function
dnorm(0,0,sqrt(2))
```

4. If $X \sim N(5,4)$, what is $P(X < 2)$? What is $P(X \leq 2)$? What is $P(X > 2)$?
```{r} 
# P(X < 2) = P(X <= 2) 
pnorm(2,5,4)

# P(X > 2) 
1-pnorm(2,5,4)
pnorm(2,5,4,lower.tail=FALSE)
```


# Poisson distribution exercises
1. Generate a random sample of size 100,000 from a Poisson distribution with rate parameter 5
```{r} 
set.seed(80417)
xsamp <- rpois(100000,5)
```

2. Compute the sample mean and variance. What do you notice?
```{r} 
mean(xsamp) 
var(xsamp)
```
3. Compute the empirical estimates of $P(X = 0)$, $P (X = 1)$, and $P(X = 2)$ for $X \sim Pois(5)$ from your sample. 
```{r, include=FALSE} 
sapply(0:2, function(x) mean(xsamp == x))

#without sapply
mean(xsamp==0)
mean(xsamp==1)
mean(xsamp==2)
```
4. Compare the estimates to the true value of the probability mass function (pmf) at $x=0,1,2$. Use the pmf and the built-in R function. Note the pmf is given by: 

$$ P(X=x) = \frac{\lambda^x e^{-\lambda}}{x!} \textrm{ where } x \in 0,1,2,... \textrm{ and } \lambda > 0$$

```{r, include=FALSE} 
#using the pmf expression
pmfpoisson <- function(x,lambda) {((lambda^x)*exp(-lambda))/factorial(x)}
pmfpoisson(0:2,5)


#using the built-in R function
dpois(0,5)
dpois(1,5)
dpois(2,5)
```

5. Compute an estimate of $P(0 < X < 3)$ from the sample and compare to true value. 
```{r} 
#estimate 
mean(xsamp > 0 & xsamp < 3)

#true value 
dpois(1,5)+dpois(2,5) #P(X=1) + P(X=2)
ppois(2,5)-ppois(0,5) #P(X <= 2) - P(X = 0)
```

6. Find the smallest $k$ such that $0.30 < P(X < k)$ for $X \sim Pois(5)$ using a while loop. Then, do this using a built-in R function. 
```{r}
# Thank you, Beau!
k = 0
pr = 0
while (pr < .3){
  pr = pr + ppois(k,5)
  k = k + 1
}
k
```


\newpage

# Group exercises
To display your results, create a table in Rmarkdown using the kable() function. Try to make it as clean as possible (i.e. column headers, title, digits, etc.). 

Fun note: you will prove these results formally in the probability course! 

## Group 1 
Generate 100,000 samples from a geometric distribution with $p = 0.3$.  Estimate $P(X \ge s+t | X \ge t)$ and $P(X \ge s)$ for  $s = 4$ and $t = 1, 2, 3, 4, 5, 6$. Compare to the true values.  What do you notice?  Google `memoryless property distribution' and take a look at the wiki page on memorylessness.  What does this suggest about the geometric distribution? 

```{r} 
xsamp <- rgeom(100000,.3)
T <- 1:6

pr_func <- function(x,s,t) {
  pr1 <- round(mean(x[x >= t] >= s + t),3)
  pr2 <- round(mean(x >= s),3)
  return(data.frame("t" = t, "cond" = pr1, "marg" = pr2))
}

results <- sapply(T,pr_func,x=xsamp,s=4)

# Create a table
kable(t(results), align=rep('c',times=3),
      col.names = c("t","P(X >= 4+t | X >= t)","P(X >= 4)"), 
      caption="Exercise 1 results")
```

## Group 2
Generate 10,000 samples from a $Bin(3, 0.5)$ and another 10,000 samples from $Bin(5, 0.5)$.  Compute the empirical cdf of the sum of the two samples and compare to the distribution function of a $Z \sim Bin(8,0.5)$ random variable.  What does this suggest about the distribution of $X+Y$ where $X \sim Bin(n_1, p)$ and $Y \sim Bin(n_2, p)$? 

```{r} 
# Two sample distribution 
x <- rbinom(10000,3,.5)
y <- rbinom(10000,5,.5)
xysum <- x+y

emp.cdf <- function(samp,val) { 
  p <- round(mean(samp <= val),3)
  return(data.frame("k" = val, "prob" = p))
}

values <- 0:8
results.sum <- sapply(values,emp.cdf,samp=xysum)

# Empirical CDF Bin(8,0.5)
results.z <- round(sapply(values,pbinom,size=8,prob=0.5),3)

# Create table
results <- cbind(t(results.sum),results.z)
kable(results,align=rep('c',times=3),
      col.names = c("k","Empirical Pr(X+Y <= k)","Pr(Z <= k)"), 
      caption="Exercise 2 results")
```


